---
title: "Wine Quality Descriptive Analysis"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Allison Jones-Farmer"
format: 
  html:
    toc: true
    toc-location: left
    toc-expand: true
    code-tools: true 
    code-fold: true
    code-overflow: scroll
    self-contained: true
---


## Data Loading and Cleaning

### Package Installation and Initialization

In this portion we use `pacman` to install the necessary packages and clear the global environment.

```{r setup,message=FALSE,warning=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE, out.width = "100%")
#package initialization
rm(list = ls()) # clear global environment
graphics.off() # close all graphics
if(require(pacman)==FALSE) install.packages("pacman")
pacman::p_load(#DataExplorer, # For explorer and cleaning data by visualizing missing values.
               gptstudio,#For coding assistance
               httr, # For downloading the data from the UCI
               tidyverse, # For data manipulation
               corrplot, # for heatmaps
               DataExplorer, #for exploring data
               skimr # for nice summary of data
               ) # For creating dummy variables of categorical variables automatically.)
```

### Load data

Download and save the data from the UCI Repository.

```{r, Read Data Files, message=FALSE}
# URLs for the wine quality data files and text file
red_wine_url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/wine-quality/winequality-red.csv"
white_wine_url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/wine-quality/winequality-white.csv"
text_file_url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/wine-quality/winequality.names"

# Directory to save the files
save_directory <- "./data/"

# Download red wine quality data file
GET(url = red_wine_url, write_disk(paste0(save_directory, "winequality-red.csv"),overwrite=T))

# Download white wine quality data file
GET(url = white_wine_url, write_disk(paste0(save_directory, "winequality-white.csv"),overwrite=T))

# Download text file
GET(url = text_file_url, write_disk(paste0(save_directory, "winequality.names"),overwrite=T))
# These lines of code remove the variables `red_wine_url`, `save_directory`,
# `text_file_url`, and `white_wine_url` from the R environment.

rm(red_wine_url)        # Remove the variable red_wine_url
rm(save_directory)       # Remove the variable save_directory
rm(text_file_url)        # Remove the variable text_file_url
rm(white_wine_url)       # Remove the variable white_wine_url
```

### Combine Files
```{r, combine files,echo=TRUE}
# Read the red wine data from the CSV file "./data/winequality-red.csv" and store it in the "red" variable
red = read.csv("./data/winequality-red.csv",sep=";")

# Read the white wine data from the CSV file "./data/winequality-red.csv" and store it in the "white" variable
white = read.csv("./data/winequality-white.csv",sep=";")

# Add a new column called "color" to the "red" data frame and set all values to "Red"
red$color = "Red"

# Add a new column called "color" to the "white" data frame and set all values to "White"
white$color = "White"

# Combine the "red" and "white" data frames using the rbind function and store the result in the "wine" variable
wine = rbind(red,white)

# Remove the "red" and "white" variables from the workspace using the rm function
rm(red)
rm(white)
```

## Overall Data Profile

```{r}
skim(wine)
```

## EDA using DataExplorer Package

Here, we shorten the path to exploration using a package with "canned" graphs.  This is handy, but the graphs are not presentation quality.  the DataExplorer package is useful for graphs as a *precursor* to your analysis.  They can give you direction.  For presentations in either a written document or a slide deck, you will most likely need to either redo the graphs in `ggplot2` or doctor these graphs up quite a bit.

### Introduce data

`plot_intro()` gives us an overview of the types of data and number of rows and columns. `plot_missing()` helps to identify missing rows in each variable.

```{r}
introduce(wine)
plot_intro(wine)
plot_missing(wine)
```

### Distributions

The functions `plot_bar()` and `plot_histogram()` create a matrix of plots for all categorical and numeric variables respectively.  They are not visually appealing graphs, but they convey the information you need.

```{r}
plot_bar(wine)
plot_histogram(wine)
```

### Boxplots by color

The `plot_boxplot` function gives side-by-side boxplots for all numeric variables. I find this helpful when trying to understand my data.

```{r}
plot_boxplot(wine, by="color")
```

### Scatterplots for Quality

The `plot_scaterplot` function gives a matrix of scatterplots for all variables.

```{r}
plot_scatterplot(wine,by="quality")
```


### Correlations

The `plot_correlations` function gives a heatmap of the correlation matrix.  It is similar to the `corrplot` function.

```{r}
plot_correlation(wine)
```

## EDA with ggplot2

### Boxplot of Qulaity

```{r}
plot <- ggplot(wine, aes(x = color, y=quality)) +
  geom_boxplot(aes(fill=color)) +
  ggtitle("Wine Quality by Color") +
  xlab("Color") +
  ylab("Quality Ratings") +
  theme_bw() 
print(plot)
```

### Correlation plot

Both wines combined

```{r}
corrplot::corrplot(cor(wine[1:12]))
```
The variables with the three highess correlations with quality are alcohol(+), volatile acidity(-), and density(+)

Lets break this down by color

```{r}
wine %>%
  filter(color=="White")%>%
  select(-color)%>%
  cor()%>%
  corrplot()
```
The three most correlated variables with quality are also alcohol(+), volatile acidity(-), and density(-).

Here is the same plot for red wine.  

```{r}
wine %>%
  filter(color=="Red")%>%
  select(-color)%>%
  cor()%>%
  corrplot()
```
The correlation patterns between quality and the most highly correlated predictors are, alcohol (+) and volatile acidity (-).

### Scatter plots of the three most correlated variables

The scatter plots show the expected bivariate relationships.  The correlations between each variable and Quality Ratings are weak, even for the three most correlated variables.

```{r}
# Alcohol
wine %>% ggplot(aes(x=alcohol, y=quality)) + 
  geom_point(color="#69b3a2",size=3) +
  xlab("Alcohol") +
  ylab("Quality Ratings") +
  ggtitle("Alcohol vs. Quality Ratings")+
  theme_bw()+
  theme(legend.position = "top")

# Volatile Acidity
wine %>% ggplot(aes(x=volatile.acidity, y=quality)) + 
  geom_point(color="#69b3a2",size=3) +
  xlab("Volatile Acidity") +
  ylab("Quality Ratings") +
  ggtitle("Volatile Acidity vs. Quality Ratings")+
  theme_bw()+
  theme(legend.position = "top")

# Sulphates
wine %>% ggplot(aes(x=density, y=quality)) + 
  geom_point(color="#69b3a2",size=3) +
  xlab("Density") +
  ylab("Quality Ratings") +
  ggtitle("Density vs. Quality Ratings")+
  theme_bw()+
  theme(legend.position = "top")
```

### Histograms of the predictors

We can see the general shapes of the distributions of wines.  We see right skewness in alcohol content, near symmetry in Volatile Acidity (red only), and symmetry in density.   

```{r}
# alcohol

wine %>% ggplot(aes(x=alcohol,fill=color))+
  geom_histogram(color="#e9ecef")+
  facet_wrap(~color)+
  xlab("Alcohol")+
  ylab("Count")+
  ggtitle("Histogram of Alcohol")+
  theme_bw()

# Volatile Acidity

wine %>% ggplot(aes(x=volatile.acidity,fill=color))+
  geom_histogram(color="#e9ecef")+
  facet_wrap(~color)+
  xlab("Volatile Acidity")+
  ylab("Count")+
  ggtitle("Histogram of Volatile Acidity")+
  theme_bw()

# Sulphates

wine %>% ggplot(aes(x=density,fill=color))+
  geom_histogram(color="#e9ecef")+
  facet_wrap(~color)+
  xlab("Density")+
  ylab("Count")+
  ggtitle("Histogram of Density")+
  theme_bw()

```
